Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-11-13 paper-bedpca.tex
save

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche, 38700, France. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

@@

\abstract{
Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses. However, conducting PCA in large populations is complicated, and there are several potential pitfalls. These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, and (3) detecting sample outliers. In this work, we explore these pitfalls of using PCA, and present efficient solutions to these pitfalls. Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.

For example, we show that PC19 to PC40 in the UK Biobank capture LD structure. Using our automatic algorithm for removing long-range LD region, we recover 16 PCs that capture population structure only.
Therefore, we recommend using only 16-18 PCs from the UK Biobank. 
We provide evidence for a shrinkage bias when projecting PCs computed with data from the 1000 Genomes project.  Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 for example suffers from a shrinkage factor of 1.50 and PC10 from a shrinkage factor of 3.14. 
We provide a fast way to project new individuals and that is not affected by this shrinkage bias. 
We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.

@@

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

@@

For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics. We make these available in a new release of R package bigsnpr (v1.0.0); this package differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on bed/bim/fam PLINK files with some missing values. 
We use them to analyse the UK Biobank data, and show that these functions are both very fast and easy to use. We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure. Interestingly, subsetting the UK Biobank data enables to get more PCs capturing population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to estimate unbiased projections instead.
Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to individuals of homogeneous ancestry.

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods.} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|c|}
\hline
Analysis & Method and/or Software & Citation & Advantages & Limitations \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
 & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assume Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

To compute this decomposition easily and efficiently, we implement an accessor that memory-map the PLINK bed file to use it directly as if it were a standard matrix. Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}. This method is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotyped matrix with a given vector, which we implement for PLINK bed files and parallelise.

@@

Finally, to choose the threshold on the previously described statistics above which to be considered as outlier, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}. 
Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster. 
In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics).

@@

To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
To project individuals from the same dataset as the ones used for computing PCA, you can use function \texttt{bed\_projectSelfPCA}. 
Note that the new individuals to be projected should not be related to the ones used for computing PCA.

@@

We provide a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
To remove related individuals with second-degree relationship or more, we apply KING-relatedness cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
To easily download these data, we provide function \texttt{download\_1000G} in R package bigsnpr.

In this paper, we also analyse the genotyped version of the UK Biobank \cite[]{bycroft2017genome}.
We apply some quality control filters; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
This results in 488,371 individuals and 504,139 variants.
When removing related individuals, we use the list of individual pairs reported by the UK Biobank.

@@

As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique. It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing statistics for detecting individual outlier samples (Section \ref{outlier-sample}).

@@

To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}).
6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic used to define outliers.\\~\\}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

When the goal is to restrict to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
Here, we choose a threshold of 5 and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
Using this threshold of 5, only 3 out of 10,936 people of Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 poeple with Black background is kept (Table \ref{tab:homogeneous}). 
In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.

@@

We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCs ($K=20$ PCs). Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}). 
When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is an apparent bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP. Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).

>>>>

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, La Tronche, 38700, France. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

@@

\abstract{
Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses. However, conducting PCA analyses can be complicated and has several potential pitfalls. These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, (3) detecting sample outliers, and (4) uneven population sizes. In this work, we explore these potential issues when using PCA, and present efficient solutions to these. Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.

For example, we show that PC19 to PC40 in the UK Biobank capture LD structure. Using our automatic algorithm for removing long-range LD regions, we recover 16 PCs that capture population structure only.
Therefore, we recommend using only 16-18 PCs from the UK Biobank. 
We provide evidence for a shrinkage bias when projecting PCs computed with data from the 1000 Genomes project.  Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 (resp.\ PC10) for example suffers from a shrinkage factor of 1.50 (resp.\ 3.14). 
We provide a fast way to project new individuals that is not affected by this shrinkage bias. 
We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.

@@

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

@@

For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics. We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on PLINK bed/bim/fam files with some missing values. 
We use these new functions to analyse the UK Biobank data, and show that they are both very fast and easy to use. We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure. Interestingly, subsetting the UK Biobank data enables to get more PCs capturing population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to estimate unbiased projections instead.
Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to individuals of homogeneous ancestry.

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods.} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|c|}
\hline
Analysis & Method and/or Software & Citation & Advantages & Limitations \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Not parallelised \\
 & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assumes Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assumes same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & Does not work for related individuals (Section \ref{proj-related}) \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

To compute this decomposition easily and efficiently, we implement an accessor that memory-map the PLINK bed file to use it directly as if it were a standard matrix. Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}. 
This algorithm is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotype matrix with a given vector, which we implement for PLINK bed files and parallelise.

@@

Finally, to choose the threshold on the previously described statistics above which variants are considered outliers, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}. 
Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their respective KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster. 
Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
We implement (the two parts of) this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics; see e.g.\ figure \ref{fig:outlier-pd}).

@@

To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
To project individuals from the same dataset as the ones used for computing PCA, you can use function \texttt{bed\_projectSelfPCA}. 
Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).

@@

We provide and use a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
To remove related individuals with second-degree relationship or more, we apply KING-relatedness cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
Resulting PLINK files and R code to generate these files are made available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
To easily download these data, we provide function \texttt{download\_1000G} in R package bigsnpr.

In this paper, we also analyse the genotyped version of the UK Biobank (\url{https://www.ukbiobank.ac.uk/}).
We apply some quality control filters; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
This results in 488,371 individuals and 504,139 variants.
When removing related individuals, we use the list of individual pairs reported by the UK Biobank.

@@

As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique. It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).

@@

To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}).
In contrast, 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
We recall that, in theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.\\~\\}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

When the goal is to restrict to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
Here, we visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
Moreover, when using this threshold, only 3 out of 10,936 people of Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 poeple with Black background is kept (Table \ref{tab:homogeneous}). 
In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.

@@

We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCs ($K=20$ PCs). Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}). 
When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is an apparent bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP. Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
\label{proj-related}
